---
title: "Exploratory Data Analysis"
author: "Grace Hyland"
format: html
editor: visual
---

## Summary Statistics

```{r}
library(tidyverse)
regular_data <- read.csv("../data/01-modified-data/skaters_clean.csv")
pp_data <- read.csv("../data/01-modified-data/skaters_pp.csv")
pk_data <- read.csv("../data/01-modified-data/skaters_pk.csv")
```

```{r}
summary_data_regular <- regular_data %>%
  subset(select=c(expectedGoals,corsi,shotAttempts,points,goals,faceoffsWon,takeways,giveaways,hits)) %>%
  summary()
summary_data_regular
```

```{r}
summary(pp_data$icetime)

#remove bottom 25% of players by ice time
pp_data <- pp_data %>%
  filter(icetime > 414)

#summarize all metrics
summary_data_pp <- pp_data %>%
  subset(select=c(expectedGoals,corsi,shotAttempts,points,goals,faceoffsWon,takeways,giveaways,hits)) %>%
  summary()
summary_data_pp
```

```{r}
summary(pk_data$icetime)

#remove bottom 25% of players by ice time
pk_data <- pk_data %>%
  filter(icetime > 555)

#summarize all metrics
summary_data_pk <- pk_data %>%
  subset(select=c(expectedGoals,corsi,shotAttempts,points,goals,faceoffsWon,takeways,giveaways,hits)) %>%
  summary()
summary_data_pk
```

## Normalization of Data

```{r}

norm_regular_data<-mutate(regular_data, expectedGoals=expectedGoals/mean(expectedGoals), corsi=corsi/mean(corsi), shotAttempts=shotAttempts/mean(shotAttempts), points=points/mean(points), goals=goals/mean(goals), faceoffsWon=faceoffsWon/mean(faceoffsWon),takeways=takeways/mean(takeways),giveaways=giveaways/mean(giveaways),hits=hits/mean(hits))

norm_pp_data<-mutate(pp_data, expectedGoals=expectedGoals/mean(expectedGoals), corsi=corsi/mean(corsi), shotAttempts=shotAttempts/mean(shotAttempts), points=points/mean(points), goals=goals/mean(goals), faceoffsWon=faceoffsWon/mean(faceoffsWon),takeways=takeways/mean(takeways),giveaways=giveaways/mean(giveaways),hits=hits/mean(hits))
 
norm_pk_data<-mutate(pk_data, expectedGoals=expectedGoals/mean(expectedGoals), corsi=corsi/mean(corsi), shotAttempts=shotAttempts/mean(shotAttempts), points=points/mean(points), goals=goals/mean(goals), faceoffsWon=faceoffsWon/mean(faceoffsWon),takeways=takeways/mean(takeways),giveaways=giveaways/mean(giveaways),hits=hits/mean(hits))
```

## Check for Outliers

```{r}
norm_regular_data<-norm_regular_data %>%
  subset(select=c(ID, name, team, expectedGoals,corsi,shotAttempts,points,goals,faceoffsWon,takeways,giveaways,hits))
sapply(norm_regular_data, function(x) sum(x > 3)) #faceoffs won has a lot of outliers

```

```{r}
norm_pp_data<-norm_pp_data %>%
  subset(select=c(ID, name, team, expectedGoals,corsi,shotAttempts,points,goals,faceoffsWon,takeways,giveaways,hits))
sapply(norm_pp_data, function(x) sum(x > 3))  #most metrics have larger number of outliers for powerplays than regular play

```

```{r}
norm_pk_data<-norm_pk_data %>%
  subset(select=c(ID, name, team, expectedGoals,corsi,shotAttempts,points,goals,faceoffsWon,takeways,giveaways,hits))
sapply(norm_pk_data, function(x) sum(x > 3))  #most metrics have larger number of outliers for penalty kills than regular play
```

## Compare Regular Play and Powerplays

```{r}
#join pp data with regular data
colnames(norm_pp_data) <- c("ID","name","team","expectedGoals_pp","corsi_pp","shotAttempts_pp","points_pp","goals_pp","faceoffsWon_pp","takeaways_pp","giveaways_pp","hits_pp")
df.1 <- left_join(norm_pp_data, norm_regular_data, by="ID")
```

```{r}
#compare metric values between regular play and powerplays for all players

library(ggplot2)

ggplot(data=df.1) + geom_point(aes(x=expectedGoals, y=expectedGoals_pp))

ggplot(data=df.1) + geom_point(aes(x=corsi, y=corsi_pp))

ggplot(data=df.1) + geom_point(aes(x=shotAttempts, y= shotAttempts_pp))

ggplot(data=df.1) + geom_point(aes(x=points, y=points_pp))

ggplot(data=df.1) + geom_point(aes(x=goals, y=goals_pp))

ggplot(data=df.1) + geom_point(aes(x=faceoffsWon, y=faceoffsWon_pp))

ggplot(data=df.1) + geom_point(aes(x=takeways, y=takeaways_pp))

ggplot(data=df.1) + geom_point(aes(x=giveaways, y=giveaways_pp))

ggplot(data=df.1) + geom_point(aes(x=hits, y=hits_pp))
```

## Compare Regular Play and Penalty Kills

```{r}
#join pk data with regular data
colnames(norm_pk_data) <- c("ID","name","team","expectedGoals_pk","corsi_pk","shotAttempts_pk","points_pk","goals_pk","faceoffsWon_pk","takeaways_pk","giveaways_pk","hits_pk")
df.2 <- left_join(norm_pk_data, norm_regular_data, by="ID")
```

```{r}
#compare metric values between regular play and penalty kills for all players

library(ggplot2)

ggplot(data=df.2) + geom_point(aes(x=expectedGoals, y=expectedGoals_pk))

ggplot(data=df.2) + geom_point(aes(x=corsi, y=corsi_pk))

ggplot(data=df.2) + geom_point(aes(x=shotAttempts, y= shotAttempts_pk))

ggplot(data=df.2) + geom_point(aes(x=points, y=points_pk))

ggplot(data=df.2) + geom_point(aes(x=goals, y=goals_pk))

ggplot(data=df.2) + geom_point(aes(x=faceoffsWon, y=faceoffsWon_pk))

ggplot(data=df.2) + geom_point(aes(x=takeways, y=takeaways_pk))

ggplot(data=df.2) + geom_point(aes(x=giveaways, y=giveaways_pk))

ggplot(data=df.2) + geom_point(aes(x=hits, y=hits_pk))
```

## Compare Powerplay and Penalty Kill Success to Team Record

-   Find powerplay and penalty kill success rates for all teams in 2022 season

-   Find data with points record for all teams in 2022 season

-   Add variable for each team that indicates whether they made the playoffs

-   Find correlation between pp and pk success and team record

```{r}
#load team data and team record data
team_data <- read.csv("../data/01-modified-data/team_record_clean.csv")

```

```{r}
#compare powerplay success to overall season success
ggplot(data=team_data) + geom_point(aes(x=ppPercent, y=points))
ggplot(data=team_data) + geom_point(aes(x=ppGoals, y=points))

#compare penalty kill success to overall season success
ggplot(data=team_data) + geom_point(aes(x=pkPercent, y=points))
ggplot(data=team_data) + geom_point(aes(x=pkGoalsAgainst, y=points))
```
